Lynott et al. (2014) - Potential error in statistics reporting

### 2018-02-07 manuscript.Rmd
minor spelling changes

abstract: >
  Psychologists must be able to test both for the presence of an effect and for the absence of an effect. In addition to testing against zero, researchers can use the Two One-Sided Tests (TOST) procedure to test for *equivalence* and reject the presence of a smallest effect size of interest (SESOI). The two-one-sided tests procedure can be used to determine if an observed effect is surprisingly small, given that a true effect at least as large as the SESOI exists. We explain a range of approaches to determine the SESOI in psychological science, and provide detailed examples of how equivalence tests should be performed and reported. Equivalence tests are an important extension of statistical tools psychologists currently use, and enable researchers to falsify predictions about the presence, and declare the absence, of meaningful effects.

@@

In any one-sided test, for an alpha level of 5%, we can reject $H_0$ when the 90% confidence interval around the observed estimate is in the predicted direction, and does not contain the value tested against (e.g., 0). In the two-one-sided tests procedure the first one-sided test is used to test against values smaller than the lower equivalence bound ($\Delta_{L}$), and the second one-sided test is used to test against values larger than the upper equivalence bound ($\Delta_{U}$). Even though the TOST procedure consists of two one-sided tests, it is not necessary to control for multiple testing because both tests need to be statistically significant to conclude equivalence. Consequently, when reporting an equivalence test, it suffices to report the one-sided test with the smaller test parameter (e.g. $t$ or $r$) and thus the larger $p$-value. Statistical equivalence can be concluded when the largest of the two $p$-values is smaller than alpha.

@@

In order to distinguish between these possibilities, we can define what a “meaningful effect” (SESOI) would be and use it as the bounds for an equivalence test. If grades for this test were set at every `r printnum(sesoi * 100)`\% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%) we could decide that we are only interested in test score differences that correspond to an increase or decrease of at least 1 grade point. Thus, our SESOI becomes a difference in raw scores of `r printnum(sesoi * 100)`\%, or `r printnum(sesoi, digits = 4)`. We can then perform an equivalence test for a two-sample Welch’s *t*-test, with equivalence bounds of $\pm$ the SESOI of `r printnum(sesoi, digits = 4)`. The two-one-sided-tests procedure consists of two one-sided tests, and we calculate *t*-values and *p*-values both for test against $\Delta_{L}$, $t(`r printnum(Moon$TOST_df)`) = `r printnum(Moon$TOST_t1)`$, $p = `r printp(Moon$TOST_p1)`$, and $\Delta_{U}$, $t(`r printnum(Moon$TOST_df)`) = `r printnum(Moon$TOST_t2)`$, $p = `r printp(Moon$TOST_p2)`$. While the *t*-test against $\Delta_{U}$ indicates that we can reject differences at least as extreme as `r sesoi`, the test against $\Delta_{L}$ shows we can *not* reject effects as large or larger than `r -sesoi`. The equivalence test is non-significant, which means we cannot reject a true effect as large or larger than 6.25% (Figure 2A), $t(`r printnum(Moon$TOST_df)`) = `r printnum(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))))`$, $p = `r printp(max(Moon$TOST_p1, Moon$TOST_p2))`$ (note we typically only report the one-sided test yielding the highest *p*-value in the result section).^[Because this is a replication of a study, it would have been reasonable to assume that we only want to reject effects in the same direction as the effect in the original study. After all, much smaller effects may indicate non-replication, but so do large effects in the opposite direction. In that case, we would perform an inferiority test (see figure 1D) against $\Delta_{U}$. As can be seen in Figure 2A, the 90\% CI does not overlap with $\Delta_{U}$, so we can reject effects as large or larger than $`r printnum(sesoi, digits = 4)`$.] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L} = -0.1$ and $\Delta_{U} = 0.1$).

>>>>

abstract: >
  Psychologists must be able to test both for the presence of an effect and for the absence of an effect. In addition to testing against zero, researchers can use the Two One-Sided Tests (TOST) procedure to test for *equivalence* and reject the presence of a smallest effect size of interest (SESOI). The two one-sided tests procedure can be used to determine if an observed effect is surprisingly small, given that a true effect at least as large as the SESOI exists. We explain a range of approaches to determine the SESOI in psychological science, and provide detailed examples of how equivalence tests should be performed and reported. Equivalence tests are an important extension of statistical tools psychologists currently use, and enable researchers to falsify predictions about the presence, and declare the absence, of meaningful effects.

@@

In any one-sided test, for an alpha level of 5%, we can reject $H_0$ when the 90% confidence interval around the observed estimate is in the predicted direction, and does not contain the value tested against (e.g., 0). In the two one-sided tests procedure the first one-sided test is used to test against values smaller than the lower equivalence bound ($\Delta_{L}$), and the second one-sided test is used to test against values larger than the upper equivalence bound ($\Delta_{U}$). Even though the TOST procedure consists of two one-sided tests, it is not necessary to control for multiple testing because both tests need to be statistically significant to conclude equivalence. Consequently, when reporting an equivalence test, it suffices to report the one-sided test with the smaller test parameter (e.g. $t$ or $r$) and thus the larger $p$-value. Statistical equivalence can be concluded when the largest of the two $p$-values is smaller than alpha.

@@

In order to distinguish between these possibilities, we can define what a “meaningful effect” (SESOI) would be and use it as the bounds for an equivalence test. If grades for this test were set at every `r printnum(sesoi * 100)`\% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%) we could decide that we are only interested in test score differences that correspond to an increase or decrease of at least 1 grade point. Thus, our SESOI becomes a difference in raw scores of `r printnum(sesoi * 100)`\%, or `r printnum(sesoi, digits = 4)`. We can then perform an equivalence test for a two-sample Welch’s *t*-test, with equivalence bounds of $\pm$ the SESOI of `r printnum(sesoi, digits = 4)`. The two one-sided-tests procedure consists of two one-sided tests, and we calculate *t*-values and *p*-values both for test against $\Delta_{L}$, $t(`r printnum(Moon$TOST_df)`) = `r printnum(Moon$TOST_t1)`$, $p = `r printp(Moon$TOST_p1)`$, and $\Delta_{U}$, $t(`r printnum(Moon$TOST_df)`) = `r printnum(Moon$TOST_t2)`$, $p = `r printp(Moon$TOST_p2)`$. While the *t*-test against $\Delta_{U}$ indicates that we can reject differences at least as extreme as `r sesoi`, the test against $\Delta_{L}$ shows we can *not* reject effects as large or larger than `r -sesoi`. The equivalence test is non-significant, which means we cannot reject a true effect as large or larger than 6.25% (Figure 2A), $t(`r printnum(Moon$TOST_df)`) = `r printnum(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))))`$, $p = `r printp(max(Moon$TOST_p1, Moon$TOST_p2))`$ (note we typically only report the one-sided test yielding the highest *p*-value in the result section).^[Because this is a replication of a study, it would have been reasonable to assume that we only want to reject effects in the same direction as the effect in the original study. After all, much smaller effects may indicate non-replication, but so do large effects in the opposite direction. In that case, we would perform an inferiority test (see figure 1D) against $\Delta_{U}$. As can be seen in Figure 2A, the 90\% CI does not overlap with $\Delta_{U}$, so we can reject effects as large or larger than $`r printnum(sesoi, digits = 4)`$.] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L} = -0.1$ and $\Delta_{U} = 0.1$).
